---
title: "Regression"
author: "Oliver Schabenberger"
date: "March 19, 2025"
output: 
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
options(width=80)
library(knitr)

knitr::opts_chunk$set(echo=TRUE, warning=F, message=F, fig.asp=0.7, cache=FALSE)

# List of required packages
required_packages <- c(
  "corrplot",
  "leaps"
)

# Function to install packages if they don't exist
install_if_missing <- function(package_name) {
  if (!requireNamespace(package_name, quietly = TRUE)) {
    cat(paste("Installing package:", package_name, "\n"))
    install.packages(package_name)
  }
  library(package_name, character.only = TRUE)
}

# Install and load all required packages
for (pkg in required_packages) {
  install_if_missing(pkg)
}
```


## Fitness

Aerobic capacity is the ability of the heart and lungs to provide the body with oxygen. It
is a measure of fitness and expressed as the oxygen intake in ml per kg body weight per
minute. Other variables that measure fitness or that are correlated with fitness are
easier to obtain, for example, the time to run a given distance or the heart rate while
running.

Measurements were made on 31 men involved in a physical fitness course at N.C.State Univ.
The variables are

-   **Age**: age in years
-   **Weight**: weight in kg
-   **Oxygen**: oxygen intake rate (ml per kg body weight per minute)
-   **RunTime**: time to run 1.5 miles (minutes)
-   **RestPulse**: heart rate while resting
-   **RunPulse**: heart rate while running (same time Oxygen rate measured)
-   **MaxPulse**: maximum heart rate recorded while running

In other words, the difficult-to-measure oxygen intake was measured along with other, more
easily attainable, variables.

### The Data

```{r fitness}
fitness <- read.csv("../data/fitness.csv")
dim(fitness)
head(fitness) 
tail(fitness)
sum(is.na(fitness))
```

The `str()` function gives us a quick view of the structure of the data frame. For each
variable in the data frame it lists the name, variable type and the first few
observations.

```{r fitness_str}
str(fitness)
summary(fitness$Weight)  
```

### Visualizations

The `pairs()` function in `R` creates a scatterplot matrix, a matrix of $(x,y)$ scatter
plots where each variable is plotted against each other variable:

```{r fitness2}
pairs(fitness)
```

You can limit the number of variables in the scatterplot matrix with the `~ x + y + ...`
syntax. The following command produces a scatterplot matrix of the inputs related to pulse
measurement.

```{r fitness3}
pairs(~ Oxygen + RestPulse + RunPulse + MaxPulse + RunTime, data=fitness)
pairs(~ Oxygen + Age + Weight, data=fitness)
```

We can customize the plot by putting histograms on the diagonal and adding a
non-parametric smoother (LOESS). The `panel.smooth` function is built in but the
`panel.hist` function needs to be added (see help for `pairs()`).

```{r fitness4}
panel.hist <- function(x, ...)
{
    usr <- par("usr")
    par(usr = c(usr[1:2], 0, 1.5) )
    h <- hist(x, plot = FALSE)
    breaks <- h$breaks 
    nB <- length(breaks)
    y <- h$counts
    y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
pairs(~ Oxygen + Weight + RunTime + RestPulse, data=fitness, 
      panel=panel.smooth, 
      diag.panel=panel.hist,
      bg="light blue")
```


```{r corr_plot}
library("corrplot")
corrplot(cor(fitness), method = "circle", diag=FALSE, tl.col="black")
```
    
### Questions

-   Is this a supervised or unsupervised analytic problem?
-   If supervised, which variable is the target (dependent variable, output) and why?
-   For the chosen target variable, how would you approach the following questions
    -   Which variables relate most strongly to the target?
    -   Which variables would you include into a model that predicts the target?
    -   How would you measure the ability of a model to predict the target for a new
        observation (an individual who did not participate in the study)?
    -   Before you apply the model to that individual, which questions would you ask?

### A linear regression

Let's perform a linear regression analysis with target `Oxygen` and `Runtime`
$$
\text{Oxygen} = \beta_0 + \beta_1 \text{Runtime}
$$

```{r fitness_lm}
linreg <- lm(Oxygen ~ RunTime, data=fitness)
summary(linreg)
```

The predicted equation is
$$
\text{Oxygen} = 82.4218 - 3.3106*\text{RunTime}
$$

```{r fit_plot}
plot(fitness$RunTime,fitness$Oxygen)
abline(coef(linreg))
```


### Multiple Linear Regression

```{r}
mlr <- lm(Oxygen ~ RunTime + RunPulse + MaxPulse, data=fitness)
summary(mlr)
```


### Variable Selection

``` {r}
library(leaps)

regfit.all <- regsubsets(Oxygen ~ ., 
                         data=fitness, 
                         method="exhaustive", 
                         nvmax=NULL)
s_all <- summary(regfit.all)
s_all

s_all$cp
s_all$bic
s_all$adjr2
```

How many models are there with 0, 1, 2, 3, ... input variables for p=6:

``` {r}
choose(6,0)
choose(6,1)
choose(6,2)
choose(6,3)
choose(6,4)
choose(6,5)
choose(6,6)
```

The total number of models you can build with p=6 input variables is

```{r}
2^6
```



```{r}
par(mfrow=c(1,3))

plot(s_all$cp, ylab="Cp", xlab="# of predictors", type="b")
idxcp <- which.min(s_all$cp) 
points(idxcp, s_all$cp[idxcp], pch=4, cex=3, col="red")

plot(s_all$bic, ylab="BIC", xlab="# of predictors", type="b")
idxbic <- which.min(s_all$bic) 
points(idxbic, s_all$bic[idxbic], pch=4, cex=3, col="red")

plot(s_all$adjr2, ylab="Adj. R^2", xlab="# of predictors", type="b")
idxr2 <- which.max(s_all$adjr2) 
points(idxr2, s_all$adjr2[idxr2], pch=4, cex=3, col="red")

mtext("All Subset Regression", side = 3, line = -2, outer = TRUE)
```

### Forward Selection

``` {r forward_selection}
regfit.forw <- regsubsets(Oxygen ~ ., 
                          data=fitness, 
                          method="forward", 
                          nvmax=NULL)
s_forw <- summary(regfit.forw)
s_forw

s_forw$cp
s_forw$bic
s_forw$adjr2
```


```{r Credit_plots_forw, echo=FALSE}
par(mfrow=c(1,3))

plot(s_forw$cp, ylab="Cp", xlab="# of predictors", type="b")
idxcp <- which.min(s_forw$cp) 
points(idxcp,s_forw$cp[idxcp], pch=4, cex=3, col="red")

plot(s_forw$bic, ylab="BIC", xlab="# of predictors", type="b")
idxbic <- which.min(s_forw$bic) 
points(idxbic,s_forw$bic[idxbic], pch=4, cex=3, col="red")

plot(s_forw$adjr2, ylab="Adj. R^2", xlab="# of predictors", type="b")
idxr2 <- which.max(s_forw$adjr2) 
points(idxr2,s_forw$adjr2[idxr2], pch=4, cex=3, col="red")

mtext("Forward Selection", side=3, line= -2, outer=TRUE)
```

### Backward Selection (Backward Elimination)

``` {r backward_selection}
regfit.back <- regsubsets(Oxygen ~ ., 
                          data=fitness, 
                          method="backward", 
                          nvmax=NULL)
s_back <- summary(regfit.back)
s_back

s_back$cp
s_back$bic
s_back$adjr2
```


```{r Credit_plots_back, echo=FALSE}
par(mfrow=c(1,3))

plot(s_back$cp, ylab="Cp", xlab="# of predictors", type="b")
idxcp <- which.min(s_back$cp) 
points(idxcp,s_back$cp[idxcp], pch=4, cex=3, col="red")

plot(s_back$bic, ylab="BIC", xlab="# of predictors", type="b")
idxbic <- which.min(s_back$bic) 
points(idxbic,s_back$bic[idxbic], pch=4, cex=3, col="red")

plot(s_back$adjr2, ylab="Adj. R^2", xlab="# of predictors", type="b")
idxr2 <- which.max(s_back$adjr2) 
points(idxr2,s_back$adjr2[idxr2], pch=4, cex=3, col="red")

mtext("Backward Elimination", side=3, line= -2, outer=TRUE)
```


### Stepwise Selection

``` {r seqrep_selection}
regfit.step <- regsubsets(Oxygen ~ ., 
                          data=fitness, 
                          method="seqrep", 
                          nvmax=NULL)
s_step <- summary(regfit.step)
s_step

s_step$cp
s_step$bic
s_step$adjr2
```

```{r Credit_plots_step, echo=FALSE}
par(mfrow=c(1,3))

plot(s_step$cp, ylab="Cp", xlab="# of predictors", type="b")
idxcp <- which.min(s_step$cp) 
points(idxcp,s_step$cp[idxcp], pch=4, cex=3, col="red")

plot(s_step$bic, ylab="BIC", xlab="# of predictors", type="b")
idxbic <- which.min(s_step$bic) 
points(idxbic,s_step$bic[idxbic], pch=4, cex=3, col="red")

plot(s_step$adjr2, ylab="Adj. R^2", xlab="# of predictors", type="b")
idxr2 <- which.max(s_step$adjr2) 
points(idxr2,s_step$adjr2[idxr2], pch=4, cex=3, col="red")

mtext("Stepwise Selection", side = 3, line = -2, outer = TRUE)
```

